Pool Model - A unified Mathematical\\ Modelling Framework for Food Science

### 2025-09-04 paper/pool_model.tex
reformulate; add more citations; fix some mistakes

In order to achieve a comprehensive description of the target system, it is important to incorporate
all relevant biological phenomena arising in these bacterial systems.
During their reproductive process, bacteria undergo multiple phases~\cite{buchanan_when_1997}, starting in a lag phase which
is necessary to adapt to new environments and initially impedes the overall growth~\cite{rolfe_lag_2012, monod_growth_1949}.
Additionally, bacteria are able to cooperate e.g. via complementary metabolic processes or inhibit
each other e.g. by means of waste production as a metabolic byproduct~\cite{west_social_2007, rainey_evolution_2003, hibbing_bacterial_2010, stubbendieck_bacterial_2016}.
A limited supply of nutrients will also affect the overall reproductive capacity of the colony~\cite{monod_growth_1949}.
Bacteria have thus devised various survival strategies to ensure sufficient access to nutrients.
Effects on a single-cell level such as adhesion~\cite{htuson_bacteriasurface_2013}, migration~\cite{decoene_microscopic_2011} and chemotaxis can lead to complex
phenomena on a population-based level which can aid in the goal of survival.
Cell signaling e.g. in the form of quorum-sensing can lead to collective emergent phenomena such as
biofilm formation which play an important role for many mature microbial communities~\cite{ng_bacterial_2009}.

@@

The first step in the construction of any mathematical model is to determine the required level of
detail with respect to the description scale and consequently which effects to include.
The bacterial growth curves let us describe the dynamics of the overall abundance of various
species.
The field of predictive microbiology has developed a variety of ODE-based models to capture
bacterial growth dynamics~\cite{perez-rodriguez_predictive_2012}.
Various \acp{ode} \textit{primary models} describe the bacterial growth curve under constant
environmental conditions.
Widely used examples include the modified Gompertz model~\cite{Gompertz1825-wi}, Baranyi-Roberts
model~\cite{baranyi_modeling_1993, baranyi_dynamic_1994} and McKellar two-compartment model~\cite{buchanan_when_1997} which distinguishes between growing and non-growing bacteria.
To take into account changing environmental conditions, so-called \textit{secondary models}
formulate the model parameters as a function of external quantities such as temperature, pH or
humidity.
These can be simple polynomial models, square root-type models~\cite{ratkowsky_relationship_1982} or
more complex relations such as the cardinal parameter model~\cite{zwietering_decision_1992}.
The advantage of the \ac{ode} formulation lies in the predictive power and simplicity of usage.
However, due to their coarse-grained nature, they lack the ability to describe heterogeneity and
other spatial effects.

@@

% To incorporate spatial heterogeneity or other effects other modeling approaches have been devised.
% \acp{pde} follow a similar pattern as \acp{ode} but also distribute the described variables through
% space.
When dealing with spatial effects such as heterogeneity, we are required to use more extensive
modeling formulations such as \acp{pde} or \acp{abm}.
The former represent an intermediate step between the population-based \ac{ode} models and the
individually-based \acp{abm} and can be effective, especially when dealing with large crowds of
cells that can be approximated as a continuum.
On the other hand, \acp{abm} excel at describing the individual behaviour of cells which can then
lead to more complex collective phenomena.
Tools such as BSim~\cite{Gorochowski2012,Dang2020} or \texttt{cellular\_raza}~\cite{Pleyer2025} are
capable to describe large amounts of cells, performing various intracellular and extracellular
processes.
The latter has been used to adapt \ac{pde}-based models into an \ac{abm} formulation which describe
bacterial branching patterns~\cite{Kawasaki1997,Matsushita1998}.

@@

In this work, we introduce a new \ac{ode}-based modeling framework which describes different
bacterial species, their metabolic states and external quantities such as nutrients as individual
pools.
The dynamics of bacterial growth can be understood in terms of transitions between the different
bacterial states i.e. pools.
Our model also allows us to capture interactions between these bacterial states and the external
microenvironment.
Compared against the aforementioned methods, this approach turns out to be less complex with general
applicability, low number of parameters and greater explainability for the underlying biological
reality.
To illustrate the generality of the model, multiple examples are reported.
We analyze the spatial limitations of an interacting two-species model and show that it can be fully
understood on an individual-based level in terms of single-cell processes.
Furthermore, we assess how spatial heterogeneity can impact the shape of the growth curves.

@@

One of the ways to describe competition between bacterial populations is as a function of the limiting resource ratio~\cite{tilman_resource_1977, smith_effects_2002}.

@@

Another mechanism is the motility of bacteria, e.g., the flagellates bacteria as \textit{E. coli} or
\textit{B. subtilis} can experience the chemotactic motion of driven by oxygen

@@

The bacterial social interactions such as releasing the toxins or public goods can also

@@

In order to construct a model which can encapsulate diffusion of nutrients and heterogeneity of
the bacteria or nutrients, we have to take into consideration the spatial component of our setup.
\acp{abm} force researchers to think in terms of cellular processes, thus bridging the gap from a

@@

We consider the system consisting from two bacterial species that can be described by two internal
states: lag and growth states.

@@

% ------------------------------------------------------------------------------------------------
\subsection{Constructing an Agent-Based Pool Model}
The following sections will demonstrate how a given \ac{ode} system can be translated into a
custom-built \ac{abm} which has only a few more overall parameters.
To construct our \ac{abm}, we use \texttt{cellular\_raza}~\cite{Pleyer2025} which was designed to
provide maximal flexibility for researchers in order to construct minimalist models.
Without the ability to define and control the precise behavior of our cell-agents, the following
approaches would not be possible.

@@

Therefore, we need to specify how many resource molecules are turned into one volumetric units $\sigma$ in \unit{\micro\metre\cubed\per\mole}.

@@

Furthermore, we let cells produce an inhibitor with production rate $b$ which inhibits the cell ability to uptake and thus use the resource by a factor of $1+\nu_I [I]$

@@

Note that we do not need to distinguish between the two species because we can set parameters individually on agent-level.
This means, we can implement the most general version of the equations and by choosing appropriate parameters, set the species of the respective cell-agent.

@@

They can be represented by their position $x_c$ and radius $r_c$ which is of course related to their (2D) volume
    V_c = \pi r_c^2

@@

Cells are not sticky and thus only pushing on each other.
No attractive force is present.
This means that colonies are expanding as individual cells grow and push each other further apart.

@@

Typically, cell cycles consist of multiple phases.

@@

Thus, we introduce an additional rule that splits apart the cell whenever a certain volume threshold is reached.
The volume of the cell is divided equally between the new and old cell and both are assigned new random positions close to each other such that their radii do touch but not overlap.

@@

    \includegraphics[width=0.49\columnwidth]{Figures/snapshot-2-cells-before-division.png}%

@@

    \includegraphics[width=0.49\columnwidth]{Figures/snapshot-2-cells-after-division.png}

@@

It is clear that this is exactly identical to the \ac{ode} model.

@@

The \ac{abm} model is more fundamental than the \ac{ode} model in the sense that it incorporates more biological aspects and maps them closer to realistic biological mechanisms.
We are thus able to derive the \ac{ode} pool model from its \ac{abm} version in the homogenous case (see supplementary material~\ref{subsec:supplement-deriving-pool-odes}).

@@

Figure~\ref{fig:spatial-snapshots} shows individual snapshots of different 

@@

    \includegraphics[width=0.49\columnwidth]{Figures/abm-homogenous/snapshot_00006000.png}\hspace{0.01\columnwidth}%
    \includegraphics[width=0.49\columnwidth]{Figures/abm-homogenous/snapshot_00012000.png}\vspace{0.01\columnwidth}\\
    \includegraphics[width=0.49\columnwidth]{Figures/abm-homogenous/snapshot_00018000.png}\hspace{0.01\columnwidth}%
    \includegraphics[width=0.49\columnwidth]{Figures/abm-homogenous/snapshot_00024000.png}%

@@

@@

We furthermore note that the total bacteria volume of the \ac{abm} model is always smaller than the \ac{ode} model.

@@

        \subcaption{Homogenous distribution, $D=30.0$\unit{\micro\metre\squared\per\minute}}

@@

        \subcaption{Homogenous distribution, $D=5.0$\unit{\micro\metre\squared\per\minute}}

@@

        \subcaption{Heterogeneous distribution, $D=30.0$\unit{\micro\metre\squared\per\minute}}

@@

        \subcaption{Randomized distribution, $D=5.0$\unit{\micro\metre\squared\per\minute}}

@@

        Comparison of the \ac{ode} and \ac{abm} model for different parameter configurations.

@@

>>>>

In order to achieve a comprehensive description of the target system, it is important to incorporate
all relevant biological phenomena arising in these bacterial systems.
During their reproductive process, bacteria undergo multiple phases~\cite{buchanan_when_1997}, starting in a lag phase which
is necessary to adapt to new environments and initially impedes the overall growth~\cite{rolfe_lag_2012, monod_growth_1949}.
Additionally, bacteria are able to cooperate e.g. via complementary metabolic processes or inhibit
each other e.g. by means of waste production as a metabolic byproduct~\cite{west_social_2007, rainey_evolution_2003, hibbing_bacterial_2010, stubbendieck_bacterial_2016}.
A limited supply of nutrients will also affect the overall reproductive capacity of the colony~\cite{monod_growth_1949}.
Bacteria have thus devised various survival strategies to ensure sufficient access to nutrients.
Effects on a single-cell level such as adhesion~\cite{htuson_bacteriasurface_2013}, migration~\cite{decoene_microscopic_2011} and chemotaxis can lead to complex
phenomena on a population-based level which can aid in the goal of survival.
Cell signaling e.g. in the form of quorum-sensing can lead to collective emergent phenomena such as
biofilm formation which play an important role for many mature microbial communities~\cite{Jin2020,ng_bacterial_2009}.

@@

The first step in the construction of any mathematical model is to determine the required level of
detail with respect to the description scale and consequently which effects to include.
The bacterial growth curves let us describe the dynamics of the overall abundance of various
species.
The field of predictive microbiology has developed a variety of ODE-based models to capture
bacterial growth dynamics~\cite{perez-rodriguez_predictive_2012}.
Various \acp{ode} \textit{primary models} describe the bacterial growth curve under constant
environmental conditions.
Widely used examples include the modified Gompertz model~\cite{Gompertz1825-wi,Gibson1987}, Baranyi-Roberts
model~\cite{baranyi_modeling_1993, baranyi_dynamic_1994} and McKellar two-compartment model~\cite{buchanan_when_1997} which distinguishes between growing and non-growing bacteria.
To take into account changing environmental conditions, so-called \textit{secondary models}
formulate the model parameters as a function of external quantities such as temperature, pH or
humidity.
These can be simple polynomial models, square root-type models~\cite{ratkowsky_relationship_1982} or
more complex relations such as the cardinal parameter model~\cite{zwietering_decision_1992}.
The advantage of the \ac{ode} formulation lies in the predictive power and simplicity of usage.
However, due to their coarse-grained nature, they lack the ability to describe heterogeneity and
other spatial effects.

@@

% To incorporate spatial heterogeneity or other effects other modeling approaches have been devised.
% \acp{pde} follow a similar pattern as \acp{ode} but also distribute the described variables through
% space.
When dealing with spatial effects such as heterogeneity, we are required to use more extensive
modeling formulations such as \acp{pde} or \acp{abm}.
The former represent an intermediate step between the population-based \ac{ode} models and the
individually-based \acp{abm} and can be effective, especially when dealing with large crowds of
cells that can be approximated as a continuum.
On the other hand, \acp{abm} excel at describing the individual behaviour of cells which can then
lead to more complex collective phenomena.
Tools such as BSim~\cite{Gorochowski2012,Dang2020}, AgentCell~\cite{emonet_agentcell_2005}, BacSim~\cite{kreft_bacsim_1998} or \texttt{cellular\_raza}~\cite{Pleyer2025} are
capable to describe large amounts of bacteria, performing various intracellular and extracellular
processes.
The latter has been used to adapt \ac{pde}-based models into an \ac{abm} formulation which describe
bacterial branching patterns~\cite{Kawasaki1997,Matsushita1998}.

@@

In this work, we introduce a new \ac{ode}-based modeling framework which describes different
bacterial species, their metabolic states and external quantities such as nutrients as individual
pools.
The dynamics of bacterial growth can be understood in terms of transitions between the different
bacterial states i.e. pools.
Our model also allows us to capture interactions between these bacterial states and the external
microenvironment.
Compared against the aforementioned methods, this approach turns out to be less complex with general
applicability, low number of parameters and greater explainability for the underlying biological
reality.
To illustrate the generality of the model, multiple examples are reported.
We analyze the spatial limitations of an interacting two-species model and show that it can be fully
understood on an individual-based level in terms of single-cell processes.
Furthermore, we assess how spatial heterogeneity~\cite{mckellar_heterogeneous_1997} can impact the shape of the growth curves.

@@

One of the ways to describe competition between bacterial populations is as a function of the limiting resource ratio~\cite{tilman_resource_1977, smith_effects_2002,ghoul_ecology_2016}.

@@

Another mechanism is the motility of bacteria, e.g., the flagellates \textit{E. coli} or
\textit{B. subtilis} which are capable of chemotactic motion of driven by oxygen

@@

The bacterial social interactions such as releas of toxins or public goods can also

@@

Compared to other modeling techniques such as \acp{pde}, \acp{abm}~\cite{nagarajan_agent-based_2022-1} force researchers to think in
terms of cellular processes, thus bridging the gap from a

@@

Our goal is to provide an \ac{abm} which can be used to derive the given \ac{ode} model as a
coarse-grained case.
In this way, we can show that the bulk-based \ac{ode} model is directly related to the mechanistic
single-cell behaviour of the bacteria.
To this end, we consider the system consisting from two bacterial species that can be described by two internal states: lag and growth states.

@@

% ------------------------------------------------------------------------------------------------
\subsection{Constructing an Agent-Based Pool Model}
The following sections will demonstrate how a given \ac{ode} system can be translated into a
custom-built \ac{abm} which has only a few more overall parameters.
This procedure is a blueprint which can be followed to construct \acp{abm} for similar
\ac{ode}-based systems.\\
To construct our \ac{abm}, we use \texttt{cellular\_raza}~\cite{Pleyer2025}.
It allows us to define our bacterial agents from a clean slate, without having to rely and inherit
existing work.
With this flexibility, researchers are able to construct minimal models using only as many parameters which are necessary.
This ability, to precisely define and control the exact behavior of our cell-agents, enables the following parts.

@@

Therefore, we need to specify how many resource molecules are turned into volumetric units $\sigma$ in \unit{\micro\metre\cubed\per\mole}.

@@

Furthermore, we let cells produce an inhibitor with production rate $b$ which inhibits the cell's ability to uptake and thus use the resource by a factor of $1+\nu_I [I]$

@@

% Note that we do not need to distinguish between the two species because we can set parameters individually on agent-level.
To distinguish between the two species, it suffices to set parameters for every individual agent accordingly.
This means, we can implement the most general version of the equations and by choosing appropriate parameters, this defines the species of the respective cell-agent.

@@

They can be represented by their position $x_c$ and radius $r_c$ which is of course related to their (2D) volume $V_c = \pi r_c^2$.

@@

We assume that the cells are not sticky and consider only a repulsive interaction potential, leaving out any possible attractive forces~\cite{Garrett2008}.
These mechanisms will result in the overall expansion of the colony as individual cells push each further other apart.

@@

Typically the cycle of a cell consist of multiple phases~\cite{Cooper2000_fn}.

@@

Thus, we introduce an additional rule that performs the division process when a certain volume threshold is reached.
The volume of the cell is divided equally between the new and old cell and both are assigned new random positions close to each other such that their radii do touch but without any overlap.

@@

    \begin{tikzonimage}[width=0.49\columnwidth]
        {Figures/snapshot-2-cells-before-division.png}%
        \node at (0.01, 0.99)[anchor=north west]{\textbf{A}};
    \end{tikzonimage}%

@@

    \begin{tikzonimage}[width=0.49\columnwidth]
        {Figures/snapshot-2-cells-after-division.png}
        \node at (0.01, 0.99)[anchor=north west]{\textbf{B}};
    \end{tikzonimage}

@@

The concatenation of many such stochastic processes on a population-based level converges identically to the lag-phase behaviour of the \ac{ode} model.

@@

The \ac{abm} model should be considered to be the more fundamental model compared to the \ac{ode} model in the sense that it incorporates more biological aspects and maps them closer to the actual behaviour of the bacteria.
We are thus able to derive the \ac{ode} pool model from its \ac{abm} version via a process of homogenization (see supplementary material~\ref{subsec:supplement-deriving-pool-odes}).

@@

Figure~\ref{fig:spatial-snapshots} shows snapshots of a numerical simulation of the described system.
The intial bacteria have been ordered on a grid such that only every diagonal neighbor is of the same species, thus resulting in a more homogenous initial distribution.

@@

    \begin{tikzonimage}[width=0.49\columnwidth]
        {Figures/abm-homogenous/snapshot_00006000.png}
        \node at (0.01, 0.99)[anchor=north west]{\textbf{A}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.49\columnwidth]
        {Figures/abm-homogenous/snapshot_00012000.png}
        \node at (0.01, 0.99)[anchor=north west]{\textbf{B}};
    \end{tikzonimage}\\
    \begin{tikzonimage}[width=0.49\columnwidth]
        {Figures/abm-homogenous/snapshot_00018000.png}
        \node at (0.01, 0.99)[anchor=north west, white]{\textbf{C}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.49\columnwidth]
        {Figures/abm-homogenous/snapshot_00024000.png}
        \node at (0.01, 0.99)[anchor=north west, white]{\textbf{D}};
    \end{tikzonimage}%

@@

Figure~\ref{fig:spatial-growth-curve} shows multiple results for various configurations of the \ac{abm}, differing only in their initial spatial configuration and diffusivity.

@@

We furthermore note that the total bacteria volume of the \ac{abm} model is always smaller or equal compared to the \ac{ode} model.

@@

        \subcaption{Hom. $D=30.0$\unit{\micro\metre\squared\per\minute}}

@@

        \subcaption{Hom. $D=5.0$\unit{\micro\metre\squared\per\minute}}

@@

        \subcaption{Het. $D=30.0$\unit{\micro\metre\squared\per\minute}}

@@

        \subcaption{Rand. $D=5.0$\unit{\micro\metre\squared\per\minute}}

@@

        Comparison of the \ac{ode} and \ac{abm} model for different configurations, showing
        homogenous (Hom.), heterogenous (Het.) or randomized (Rand.) initial configurations.

@@

These results show how variations in properties that can not be described by simple \ac{ode} models can lead to large deviations in the growth curves of the species.
Furthermore, it was shown that the utilized pool model can be derived from first principles by performing a homogenization of the mechanistic bacterial properties.
